---
title: "Intro to R"
author: "Jacob Carey"
date: \today
header-includes:
   - \usepackage{inconsolata}
output: 
  beamer_presentation:
    toc: true
    theme: "Szeged"
    slide_level: 2
---

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.height=3, fig.height=3)

# matrix example
library(CoGAPS)
data(SimpSim)
micro.array.data <- SimpSim.D[1:5, 1:5]

# data frame example
set.seed(1)
case.control <- data.frame(case.status=rbinom(10, 1, 0.5),
                           father.status=sample(c("Diseased",
                                                  "Not diseased"),
                                                2))
```

# Why R?

## Data Structures

- R is a *statistical* **programming** language
- Many provided *data structures* for handling distinct scenarios
- A matrix can handle $n$ rows and $m$ columns of numeric data
    - Use case: microarray data for $n$ genes from $m$ samples
- Data frames are a robust solution to handle all sorts of tabular data
    - Use case: case control status of individuals with candidate heritable traits of parents listed

## Example (Matrix)

```{r}
# Microarray data
micro.array.data
```

## Example (Data Frame)

```{r}
# Microarray data
case.control
```

## Bioconductor

- Repository for third party R code
- "Bioconductor provides tools for the analysis and comprehension of high-throughput genomic data."

# Basics

## Adding Numbers and Calculating Tips

```{r}
# What's 2 + 2?
2 + 2

# How much do I need to tip on $40?
40 * 0.15

# What's the probability of three heads?
0.5 ^ 3
```

## Variables

- Variables (among other things) separate R from a calculator
    - The `<-` symbol indicates **assignment** of a *value* to *variable*
- Allow you to *descriptively* name an important number for combination with other *descriptively* named numbers
- Lots of restrictions about them

- There are four primary types of variables
    1. Numeric (1.45)
    2. Integer (3)
    3. Character ("hello, world")
    4. Boolean (TRUE)

- Naming variables
    - contains only characters (case matters!!), numbers, periods, and underscores
    - Must begin with a character
    - Be descriptive!

## Variables

```{r, error=TRUE}
variable.number <- 4 # store a number in a variable
variable.number

variable.character <- "hello" # store a word
variable.number * 2 # use the variable to do something

# can I add a number and a character?
variable.number + variable.character 
```

## Vectors

- Vectors in R a (ordered, but possibly not sorted) collection of *one type of variable*
- They allow you to perform operations across multiple numbers, characters, etc
    - Allow you to do stuff with basic *data*
- One way to create a variable is through the `c` *command*
- Additionally, R supports the `:` operator to generate a sequence of integers

## Vectors

```{r}
vector.simple <- 1:5
vector.simple # inspect contents
vector.numbers <- c(4.3, 7.1, 5.0, 3.1) # numerics
vector.error <- c("ab", 4.3, 2.1, 17.8)
vector.error # inspect contents
```

- notice that `vector.error` has *coerced* the numeric data to character data
    - Remember: a vector is a collection of **one type of variable**

## Comparisons

- R supports comparison operators
    - `>` greater than
    - `<` less than
    - `==` Equal to
    - `!=` Not equal to
- For vector data, comparisons are made element-wise 

## Comparisons

```{r}
4 > 2
vector.numbers == 7.1
vector.numbers < 6
```

## Functions

- R allows you to use many *built-in* as well as *third-party* (available through downloadable packages) to do many useful things with data
- Want to calculate the mean of your data? Use a function!
- Want to do a t-test? Use a function!
- Functions are syntactically written as `function.name(arg1, arg2, ...)` 
    - Number of arguments can vary from 0 to essentially unlimited
    - Arguments can be variables, vectors, etc. 
    - Arguments can sometimes be other functions!

## Functions

```{r}
homework <- c(99, 85, 90, 86, 30, 75)
mean(homework) # what's the mean homework score?
sd(homework) # what's the standard deviation of homeworks?
```

## Quiz

\begin{columns}
\begin{column}{0.48\textwidth}
1. Which of these expressions will give an error when evaluated in R? \\
    a. \texttt{2 >= 2} \\
    b. \texttt{x=245} \\
    c. \texttt{3=3} \\
    d. \texttt{17==17} \\
2. What type of variable is x, if I assign it like this:  x <- "6" \\
    a. data frame \\
    b. integer \\
    c. numeric \\
    d. character
\end{column}
\begin{column}{0.48\textwidth}
3. My code: \\
   \texttt{x <- 30} \\
   \texttt{y <- 0.6} \\
   \texttt{z <- x+y} \\
   \texttt{y <- 17} \\
What is the value z? \\
    a. 30 \\
    b. 47 \\
    c. 30.6 \\
    d. 17.6
\end{column}
\end{columns}

# Moving beyond calculators

## Reading in Data

- Data can be read in from a local file (such as a CSV)
- Similarly, data be read in from a url
    - Both ways just require 1) a function and 2) a character revealing the location of the file
- Simplest function for reading CSV data is `read.csv`

```{r}
site <- "http://www.ats.ucla.edu/stat/data/hsb2.csv"
# same function for reading local data
ucla.data <- read.csv(site) 
```

## Looking at data

```{r}
head(ucla.data) # inspect data using the head function
ucla.data[100, ] # view the 100th row
```

## Looking at data

```{r}
summary(ucla.data) # calculate default summary statistics of every variable
```

## Data Frames

- What type of variable is `ucla.data`?
    - It's a **data frame** which is just a collection of *named* vectors that are all the same length
    - These individual *variables* or *columns* can be accessed using the cryptic looking `$` symbol

## Data Frames

```{r}
# t test comparing the math and science distributions
t.test(ucla.data$math, ucla.data$science, paired=TRUE)
```

## Data Frames

- Word of *warning*
    - The `attach` function can be used to access variables in a dataset without the dataset name and `$`, but can cause issues further down the line
    - Avoid `attach` if you can!
- Instead, many functions support `data` as an argument, and allow variables to be called by name

## Data Frames

```{r}
# are the writing scores between genders different?
t.test(write ~ female, data=ucla.data)
```

## Data Frames

```{r}
# 'subset' the data and only keep the female students
female.data <- subset(ucla.data, # data name
                      # conditions for keeping data 
                      # (note the '==' equality)
                      female == 1)
```

## Data Frames

```{r}
# keep the second column
column.two <- ucla.data[, 2]

# keep all columns and rows 4 to 6
rows.four.to.six <- ucla.data[4:6, ]

# keep columns one and three and rows 2 through 5
subsetted.data <- ucla.data[2:5, c(1, 3)]
```

Note that these same subset methods will also work with matrix data!

## Installing packages

Simple!

```{r, eval=FALSE}
install.packages("ggplot2") # install the package
library(ggplot2)            # now load the package
```

# Basic statistics with R

## Correlation, Covariance, and Variance

```{r}
var(ucla.data$write)
cov(ucla.data$write, ucla.data$read)
cor(ucla.data$write, ucla.data$read)
```

## Regression

```{r} 
# make inference about math from science and reading 
# (linear regression)
lm(math ~ science + read, data=ucla.data)
```

## Regression

```{r}
# inference about gender from reading (logistic)
glm(female ~ science + read, data=ucla.data, 
    family=binomial)
```

## Random Variables in R

>- Many times we have assumptions about the data and models, but program the analysis incorrectly
>- How can we check for this?
>- R allows for drawing from parameterized distributions!

## Random Variables in R

Basic scenario: 50 observations from the same normal distribution and we are interested in the mean

```{r}
y <- rnorm(n=50,    # number of draws
           mean=14, # mean of distribution
           sd=5)    # Standard deviation
mean(y)
```

Many of other provided distributions, such as binomial, gamma, beta, etc.

# Plotting

## Intro

- Analysis is all well and good, but how do we *look* at our data?
- Many options but most people use either
    1. Base R functions
    2. The ggplot2 package (Not discussed here)

## Histograms
```{r}
# histogram
hist(ucla.data$write)
```

## Scatterplot

```{r, eval=FALSE}
# note that functions can run across multiple lines!
plot(ucla.data$read, ucla.data$write, # x and y data
     main="Plot of scores", # main title
     xlab="Reading", # x axis label
     ylab="Writing") # y axis label

# add a line comparing math and science
lines(sort(ucla.data$math),    # functions inside of functions!
      sort(ucla.data$science), 
      col="blue")              # color this line blue

# add a smoother to the scatter plot
lines(lowess(ucla.data$read, ucla.data$write),
      col="orange")            # color this line orange
```

## Scatterplot

```{r, echo=FALSE, fig.height=7, fig.width=7}
plot(ucla.data$read, ucla.data$write, # x and y data
     main="Plot of scores", # main title
     xlab="Reading", # x axis label
     ylab="Writing") # y axis label

# add a line comparing math and science
lines(sort(ucla.data$math),    # functions 
      sort(ucla.data$science), # inside of functions!
      col="blue")              # color this line blue

# add a smoother to the scatter plot
lines(lowess(ucla.data$read, ucla.data$write),
      col="orange")           # color this line orange
```

# Further work / Advanced

## Getting Help?

**Switch to RStudio**

## if statements

`if` statements create conditional outputs

```{r}
x <- 4

if (x > 6) {
    print("x is greater than 6") 
} else if (x <= 6 && x > 0) {
    print("x is less than or equal to 6 and greater than 0") 
} else {
    print("x is less than or equal to 0")
}
```

## Custom functions 

```{r, eval=FALSE}
function.name <- function(arg1, arg2) {
    args.added <- arg1 + arg2

    return(args.added)
}
```

## Custom functions 

```{r}
pointwise <- function(x, limit=0) {
    if (x < limit) {
        return(0) 
    } else {
        return(x)
    }
}

pointwise(4)
```

## Custom functions 

- Note about "optional arguments"
    - If an argument has an associated value in function construction, that argument is optional and will take the listed default value if not passed to the function

## For loops

Combined with custom functions and you can do pretty much anything you want!

```{r}
collection <- 1:5
for (x in collection) {
    print(x ^ 2)
}
```

## Questions?
