---
title: "Intro to R"
author: "Jacob Carey"
date: \today
output: 
  beamer_presentation:
    theme: "Szeged"
    slide_level: 2
---

# Basics

## Adding Numbers and Calculating Tips

```{r}
# What's 2 + 2?
2 + 2

# How much do I need to tip on $40?
40 * 0.15

# What's the probability of three heads?
0.5 ^ 3
```

## Variables

- Variables (among other things) separate R from a calculator
- Allow you to *descriptively* name an important number for combination with other *descriptively* named numbers
- Lots of restrictions about them

- There are four primary types of variables
    1. Numeric (1.45)
    2. Integer (3)
    3. Character ("hello, world")
    4. Boolean (TRUE)

- Naming variables
    - contains only characters (case matters!!), numbers, periods, and underscores
    - Must begin with a character
    - Be descriptive!

```{r, error=TRUE}
variable.number <- 4 # store a number in a variable
variable.number

variable.character <- "hello" # store a word in a variable

variable.number * 2 # use the variable to do something

variable.number + variable.character # can I add a number and a character?
```

# Actually being useful

## Reading in Data

## Getting Help?


