---
title: "Intro to R"
author: "Jacob Carey"
date: \today
output: 
  beamer_presentation:
    toc: true
    theme: "Szeged"
    slide_level: 2
---

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.height=3, fig.height=3)
```

# Basics

## Adding Numbers and Calculating Tips

```{r}
# What's 2 + 2?
2 + 2

# How much do I need to tip on $40?
40 * 0.15

# What's the probability of three heads?
0.5 ^ 3
```

## Variables

- Variables (among other things) separate R from a calculator
    - The `<-` symbol indicates **assignment** of a *value* to *variable*
- Allow you to *descriptively* name an important number for combination with other *descriptively* named numbers
- Lots of restrictions about them

- There are four primary types of variables
    1. Numeric (1.45)
    2. Integer (3)
    3. Character ("hello, world")
    4. Boolean (TRUE)

- Naming variables
    - contains only characters (case matters!!), numbers, periods, and underscores
    - Must begin with a character
    - Be descriptive!

## Variables

```{r, error=TRUE}
variable.number <- 4 # store a number in a variable
variable.number

variable.character <- "hello" # store a word
variable.number * 2 # use the variable to do something

# can I add a number and a character?
variable.number + variable.character 
```

## Vectors

- Vectors in R a (ordered, but possibly not sorted) collection of *one type of variable*
- They allow you to perform operations across multiple numbers, characters, etc
    - Allow you to do stuff with basic *data*
- One way to create a variable is through the `c` *command*
- Additionally, R supports the `:` operator to generate a sequence of integers

## Vectors

```{r}
vector.simple <- 1:5
vector.simple # inspect contents
vector.numbers <- c(4.3, 7.1, 5.0, 3.1) # numerics
vector.error <- c("ab", 4.3, 2.1, 17.8)
vector.error # inspect contents
```

- notice that `vector.error` has *coerced* the numeric data to character data
    - Remember: a vector is a collection of **one type of variable**

## Comparisons

- R supports comparison operators
    - `>` greater than
    - `<` less than
    - `==` Equal to
    - `!=` Not equal to
- For vector data, comparisons are made element-wise 

## Comparisons

```{r}
4 > 2
vector.numbers == 7.1
vector.numbers < 6
```

## Functions

- R allows you to use many *built-in* as well as *third-party* (available through downloadable packages) to do many useful things with data
- Want to calculate the mean of your data? Use a function!
- Want to do a t-test? Use a function!
- Functions are syntactically written as `function.name(arg1, arg2, ...)` 
    - Number of arguments can vary from 0 to essentially unlimited
    - Arguments can be variables, vectors, etc. 
    - Arguments can sometimes be other functions!

## Functions

```{r}
homework <- c(99, 85, 90, 86, 30, 75)
mean(homework) # what's the mean homework score?
sd(homework) # what's the standard deviation of homeworks?
```

# Moving beyond calculators

## Reading in Data

- Data can be read in from a local file (such as a CSV)
- Similarly, data be read in from a url
    - Both ways just require 1) a function and 2) a character revealing the location of the file
- Simplest function for reading CSV data is `read.csv`

```{r}
site <- "http://www.ats.ucla.edu/stat/data/hsb2.csv"
ucla.data <- read.csv(site) # same function for reading local data
```

## Looking at data

```{r}
head(ucla.data) # inspect data using the head function
ucla.data[100, ] # view the 100th row
```

## Looking at data

```{r}
summary(ucla.data) # calculate default summary statistics of every variable
```

## Data Frames

- What type of variable is `ucla.data`?
    - It's a **data frame** which is just a collection of *named* vectors that are all the same length
    - These individual *variables* or *columns* can be accessed using the cryptic looking `$` symbol

## Data Frames

```{r}
# t test comparing the math and science distributions
t.test(ucla.data$math, ucla.data$science, paired=TRUE)
```

## Data Frames

- Word of *warning*
    - The `attach` function can be used to access variables in a dataset without the dataset name and `$`, but can cause issues further down the line
    - Avoid `attach` if you can!
- Instead, many functions support `data` as an argument, and allow variables to be called by name

## Data Frames

```{r}
# are the writing scores between genders different?
t.test(write ~ female, data=ucla.data)

# 'subset' the data and only keep the female students
female.data <- subset(ucla.data, # data name
                      female == 1) # conditions for keeping data (note the '==' equality)
```

## Installing packages

Simple!

```{r, eval=FALSE}
install.packages("ggplot2") # install the package
library(ggplot2)            # now load the package
```

# Basic statistics with R

## Correlation, Covariance, and Variance

```{r}
var(ucla.data$write)
cov(ucla.data$write, ucla.data$read)
cor(ucla.data$write, ucla.data$read)
```

## Regression

```{r} 
# make inference about math from science and reading 
# (linear regression)
lm(math ~ science + read, data=ucla.data)
```

## Regression

```{r}
# inference about gender from reading (logistic)
glm(female ~ science + read, data=ucla.data, 
    family=binomial)
```

# Plotting

## Intro

- Analysis is all well and good, but how do we *look* at our data?
- Many options but most people use either
    1. Base R functions
    2. The ggplot2 package

## Base R
```{r}
# histogram
hist(ucla.data$write)
```

## Base R

```{r, eval=FALSE}
# note that functions can run across multiple lines!
plot(ucla.data$read, ucla.data$write, # x and y data
     main="Plot of scores", # main title
     xlab="Reading", # x axis label
     ylab="Writing") # y axis label

# add a line comparing math and science
lines(sort(ucla.data$math),    # functions inside of functions!
      sort(ucla.data$science), 
      col="blue")              # color this line blue

# add a smoother to the scatter plot
lines(lowess(ucla.data$read, ucla.data$write),
      col="orange")            # color this line orange
```

## Base R

```{r, echo=FALSE, fig.height=7, fig.width=7}
plot(ucla.data$read, ucla.data$write, # x and y data
     main="Plot of scores", # main title
     xlab="Reading", # x axis label
     ylab="Writing") # y axis label

# add a line comparing math and science
lines(sort(ucla.data$math),    # functions 
      sort(ucla.data$science), # inside of functions!
      col="blue")              # color this line blue

# add a smoother to the scatter plot
lines(lowess(ucla.data$read, ucla.data$write),
      col="orange")           # color this line orange
```

## ggplot2

ggplot2 graphing requires very different syntax from base R

```{r, message=FALSE}
# note that every new session of R requires 
# packages to be loaded manually
library(ggplot2)
```

## ggplot2

```{r, message=FALSE}
# histogram
ggplot(data=ucla.data, aes(x=write)) +
    geom_histogram() + 
    xlab("Writing") +
    ggtitle("Distribution")
```

## ggplot2

```{r, message=FALSE}
# investigate distribution of science scores by SES
ggplot(data=ucla.data, aes(x=science)) +
    geom_histogram(aes(fill=as.character(ses))) + 
    # note that grouping variable must be a character
    xlab("Writing") + 
    ggtitle("Distributions by Socioeconomic Status")
```

# Further work / Advanced

## Getting Help?

**Switch to RStudio**

## if statements

`if` statements create conditional outputs

```{r}
x <- 4

if (x > 6) {
    print("x is greater than 6") 
} else if (x <= 6 && x > 0) {
    print("x is less than or equal to 6 and greater than 0") 
} else {
    print("x is less than or equal to 0")
}
```

## Custom functions 

```{r, eval=FALSE}
function.name <- function(arg1, arg2) {
    args.added <- arg1 + arg2

    return(args.added)
}
```

## Custom functions 

```{r}
pointwise <- function(x, limit=0) {
    if (x < limit) {
        return(0) 
    } else {
        return(x)
    }
}

pointwise(4)
```

## Custom functions 

- Note about "optional arguments"
    - If an argument has an associated value in function construction, that argument is optional and will take the listed default value if not passed to the function

## For loops

Combined with custom functions and you can do pretty much anything you want!

```{r}
collection <- 1:5
for (x in collection) {
    print(x ^ 2)
}
```

## Questions?
