---
title: "Intro to R"
author: "Jacob Carey"
date: \today
output: 
  html_document
#  beamer_presentation:
#    theme: "Szeged"
#    slide_level: 2
---

# Basics

## Adding Numbers and Calculating Tips

```{r}
# What's 2 + 2?
2 + 2

# How much do I need to tip on $40?
40 * 0.15

# What's the probability of three heads?
0.5 ^ 3
```

## Variables

- Variables (among other things) separate R from a calculator
- Allow you to *descriptively* name an important number for combination with other *descriptively* named numbers
- Lots of restrictions about them

- There are four primary types of variables
    1. Numeric (1.45)
    2. Integer (3)
    3. Character ("hello, world")
    4. Boolean (TRUE)

- Naming variables
    - contains only characters (case matters!!), numbers, periods, and underscores
    - Must begin with a character
    - Be descriptive!

```{r, error=TRUE}
variable.number <- 4 # store a number in a variable
variable.number

variable.character <- "hello" # store a word in a variable

variable.number * 2 # use the variable to do something

variable.number + variable.character # can I add a number and a character?
```

## Vectors

- Vectors in R a (ordered, but possibly not sorted) collection of *one type of variable*
- They allow you to perform operations across multiple numbers, characters, etc
    - Allow you to do stuff with basic *data*
- One way to create a variable is through the `c` *command*

```{r}
vecture.numbers <- c(4.3, 7.1, 5.0, 3.1) # collection of numerics
vector.error <- c("ab", 4.3, 2.1, 17.8) # must be one type of variable!
vector.error # inspect contents
```

- notice that `vector.error` has *coerced* the numeric data to character data
    - Remember: a vector is a collection of **one type of variable**

## Functions

- R allows you to use many *built-in* as well as *third-party* (available through downloadable packages) to do many useful things with data
- Want to calculate the mean of your data? Use a function!
- Want to do a t-test? Use a function!
- Functions are syntactically written as `function.name(arg1, arg2, ...)` 
    - Number of arguments can vary from 0 to essentially unlimited
    - Arguments can be variables, vectors, etc. 
    - Arguments can sometimes be other functions!

```{r}
homework <- c(99, 85, 90, 86, 30, 75) # c is a function too!
mean(homework) # what's the mean homework score?
sd(homework) # what's the standard deviation of the homework score?
```

# Actually being useful

## Reading in Data

- Data can be read in from a local file (such as a CSV)
- Similarly, data be read in from a url
    - Both ways just require 1) a function and 2) a character revealing the location of the file
- Simplest function for reading CSV data is `read.csv`

```{r}
ucla.data <- read.csv("http://www.ats.ucla.edu/stat/data/hsb2.csv") # same function for reading local data
head(ucla.data) # inspect data
```

## Getting Help?


